---
title: "Exploring Length, Number, and Types of Felony Cases Sentenced in Cook County"
author: "Damini Sharma"
date: "1/27/2019"
output: 
  html_document:
    code_folding: hide

---

```{r warning=FALSE , message=FALSE, include=FALSE}
# load necessary packages 
libs <- c("tidyverse", "stringr", "readr", "dplyr", "ggplot2", "readstata13","foreign",
          "magrittr","lubridate","here","ggrepel")
lapply(libs, library, character.only=TRUE)


```

The Cook County State's Attorney's Office (SAO) is the second largest prosecutor's office in the United States. Tens of thousands of criminal cases are prosecuted every year by the Office, across 100 law enforcement jurisdictions and in service of over five million residents of Cook County. When she took office in December 2016, State's Attorney Kim Foxx announced a largely unprecedented commitment to improving the transparency of the SAO's work. Accordingly, in 2017 the SAO released a Baseline Data Report, along with de-identified data capturing snapshots of all cases that go through the SAO from arrest to sentencing. While some of the data goes as far back as the early 2000s, due to changes in the accuracy of storing and reporting the data, it is more reliable starting in 2011. 

Cook County has been criticized of having excessive lags in the process from arrest to sentencing. Such long delays can prove incredibly disruptive in the life of a defendant, especially if they are held in jail. Using the data released by the SAO, as well as additional open datasets, I would like to see if the data does, indeed, show long lags between incident, arrest, felony review, arraignment, and sentencing. In particular, I’d be interested to see whether these lags have any variation when looked at across time, types of offenses, districts, or judges. Understanding these nuances could shed light on where the bottleneck is happening, if at all.

I start by looking at the Sentences data from the SAO to see how the average length of case (from arrest to sentencing) has changed from 2011 to 2018. In addition to the average case length, I plot total number of defendants that are sentenced and the total number of days spent on cases in the same time period. All three are plotted as percentage of the February 2011 value.


```{r warning=FALSE, message = FALSE}
# Read in the data
intake <- read.csv(here("data_source","SAO Data","Intake.csv"),stringsAsFactors = FALSE) 
initiation <- read.csv(here("data_source","SAO Data","Initiation.csv"),stringsAsFactors = FALSE) 
sentence <- read.csv(here("data_source","SAO Data","Sentencing.csv"),stringsAsFactors = FALSE) 
disposition <- read.csv(here("data_source","SAO Data","Dispositions.csv"),stringsAsFactors = FALSE) 

# Clean data

# For graph 1, we look at sentences data only. 
# There is already a length of case in days variable but this is from arraignment to 
# sentencing, and I want to start at arrest so I define my own

# Since the sentencing data is not unique on case participant ID, I first collapse the data
# to get an average case length per participant ID. This is because many defendants will have
# multiple charges, and say each charge takes 8 days to go from arraignment to sentencing. Those
# are the same 8 days, so we don't want to double count them.
sentence %<>%
  mutate(SENTENCE_DATE = as.Date(SENTENCE_DATE,'%m/%d/%Y' )) %>% 
  mutate(ARREST_DATE = as.Date(ARREST_DATE,'%m/%d/%Y' )) %>% 
  mutate(case_length_total = unclass(difftime(SENTENCE_DATE,ARREST_DATE,units = "days"))) %>%
  mutate(sentence_month = round_date(SENTENCE_DATE, "month"))  

case_length <- sentence %>% 
  group_by(sentence_month, CASE_PARTICIPANT_ID) %>% 
  summarise(case_length_p = mean(case_length_total, na.rm = TRUE)) %>% 
  group_by(sentence_month) %>% 
  summarise(felonies_total = n_distinct(CASE_PARTICIPANT_ID), 
            case_length_total = sum(case_length_p, na.rm = TRUE),
            avg_case_length = mean(case_length_p, na.rm = TRUE)) %>% 
  filter(felonies_total > 0) %>% 
  filter(case_length_total > 0) %>% 
  filter(avg_case_length > 0 & avg_case_length < 2000) %>% 
  filter(sentence_month > "2011-01-01" & sentence_month < "2018-01-01") %>% 
  gather(type_measure,n,felonies_total:avg_case_length)

case_length_2011 <- case_length %>% filter(sentence_month == "2011-02-01") %>% select(-sentence_month)

# merge this value back onto long dataset
case_length_as_pct <- left_join(case_length, case_length_2011, by="type_measure") %>% 
  mutate(pct_change = (n.x / n.y))


plot1 <- 
  case_length_as_pct %>% 
  ggplot(aes(x=sentence_month, y=pct_change, color=type_measure)) + 
  geom_point(size = 0.8) +
  geom_smooth(size=0.3) +
  scale_color_manual(values=c("chocolate4","gray60","darkorange")) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0.8 , 1.9)) +
  labs(
    subtitle = "Monthly Number of Defendants Sentenced, Total Days Spent on Cases, and Average Case \n Length as a Percent of February 2011 Value",
    title = "Average Case Length Increases Even As Total Defendants \n Sentenced Decreases",
    y="",
    x = "Year",
    caption = "Source: Cook County State Attorney's Office Data Portal - Sentences"
  ) +
  geom_text(aes(x = as.Date("2015-06-01"), y = 1.75, label = "Total Number of Days \n Spent on Cases"),color="gray60") +
  geom_text(aes(x = as.Date("2015-05-01"), y = .90, label = "Total Number of \n Defendants Sentenced"),color="darkorange") +
  geom_text(aes(x = as.Date("2017-09-01"), y = 1.30, label = "Average \n Case Length"),color="chocolate4") +
  geom_hline(yintercept = 1, color = "chocolate2", linetype="dashed")

ggsave(here("data_output","cases_sentenced_avg_case_length.pdf"),plot1,device="pdf",width=8,height=5)
plot1
```

Average case length increases consistently from 2011 to 2017. Until approximately 2015, the number of defendants sentenced per month is rising and total number of days is rising at a faster rate which could explain the rise in average case length from 2011 to 2015. As the number of defendants increases, the Office is under more pressure, which would cause the total number of days spent to increase at a faster rate. However, if this was the full explanation, we would see the total number of days decrease as the number of defendants declines sharply after 2015. Instead, we see that the total number of days actually increase from 2015 to 2017. From 2015 to 2017, there seem to be two simultaneous trends that are increasing the average case length: decreasing number of defendants $\textit{and}$ an increasing number of days spent on cases. In other words, the Office is spending more days on fewer cases.

To explain this trend further, I turn to types of offenses. I want to understand which offense types are driving the decline in number of defendants, as well as which offense types are driving the rise in total number of days spent. In order to do this, I first look at the breakdown of offense types for all cases sentenced from 2011 to 2017.

```{r warning=FALSE, message = FALSE}
intake %<>% 
  mutate(RECEIVED_DATE = as.Date(RECEIVED_DATE,'%m/%d/%Y')) %>% 
  mutate(receive_month = round_date(RECEIVED_DATE, "month")) %>% 
  mutate(Offense_Category_broad = case_when(
    grepl("Retail Theft", Offense_Category) ~ "Retail Theft",
    grepl("Burglary", Offense_Category) ~ "Burglary",
    grepl("Homicide", Offense_Category) ~ "Homicide",
    grepl("Robbery", Offense_Category) ~ "Robbery",
    grepl("Battery", Offense_Category) ~ "Battery",
    grepl("Assault", Offense_Category) ~ "Battery",
    grepl("DUI", Offense_Category) ~ "DUI",
    grepl("UUW", Offense_Category) ~ "Unlawful Use of Weapon",
    Offense_Category == "Theft" ~ "Theft",
    Offense_Category == "Narcotics" ~ "Narcotics",
    Offense_Category == "Sex Crimes" ~ "Sex Crimes",
    Offense_Category == "Possession of Stolen Motor Vehicle" ~ "MVT",
    Offense_Category == "Driving With Suspended Or Revoked License" ~ "Driving With \n Revoked License",
    TRUE ~ "Other (e.g. Forgery, \n Identity Theft)"    
  )) 

intake_other <- intake %>% 
  filter(Offense_Category_broad == "Other (e.g. Forgery, \n Identity Theft)")  %>% 
  count(Offense_Category)

intake_initiation <- inner_join(intake,initiation, by = "CASE_PARTICIPANT_ID")
int_ini_sent <- inner_join(intake_initiation,sentence,by="CASE_PARTICIPANT_ID") 

intake_count <- int_ini_sent %>% 
  mutate(year = round_date(receive_month,"year")) %>% 
  filter(year < "2018-01-01" & year >= "2011-01-01") %>% 
  group_by(Offense_Category_broad) %>% 
  summarise(felonies = n_distinct(CASE_PARTICIPANT_ID)) %>% 
  mutate(pct = round(felonies / sum(felonies),4)) 
  
plot2<- intake_count %>% 
  ggplot(aes(x=reorder(Offense_Category_broad,-pct), y = pct)) +
  geom_point(size=3, color = "chocolate4") +
  geom_segment(aes(x=Offense_Category_broad,
                  xend=Offense_Category_broad,
                  y = 0,
                  yend = pct)) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0.0,.45),breaks = seq(0,.45,.05)) +
  scale_x_discrete(position = "top") +
  theme_minimal() +
  #theme(axis.title.y=element_blank(),
  #      axis.text.y=element_blank(),
  #      axis.ticks.y=element_blank()) +
  theme(panel.grid.minor = element_blank()) +
labs(
    title = "Narcotics, Retail Theft, DUIs, and Unlawful Use of Weapon \n Make Up 60% of All Offenses Sentenced",
    subtitle = "Felonies Sentenced from 2011 to 2017",
    y="Proportion of Offenses Sentenced",
    x = "",
    caption = "Source: Cook County State Attorney's Office Data Portal - Intake, Initiations, and Sentences"
  ) 

ggsave(here("data_output","breakdown_offense_type.pdf"),plot2,device="pdf",width=8,height=5)
plot2

```

It should be noted here that defendants can have multiple offense types. For example, someone arrested for burglarly may also have been found with drugs on them, and may be prosecuted and sentenced for both charges separately. Instead of using offense types at the charge level, I use the broad "Offense Type" variable from the Intake file, which represents all case participants -- not charges -- that are brought in to the SAO for review. If the SAO decides to prosecute, a case is "initiated" and the specific charges are determined. To simplify the analysis, I use the broad offense type from the Intake file and merge this onto the sentences data. 

Narcotics, at approximately 30\%, make up the largest portion of all offenses sentenced. Collectively, narcotics, unlawful use of weapon, DUIs, retail theft, burglary, and battery make up approximately 70\% of all offenses sentenced. The "Other Crimes" category contains approximately 60 distinct crimes such as forgery, escape (failure to return), identity theft and credit card fraud. 

Next, I want to look at which of these offenses are, over time, contributing to the trend of decreasing number of cases we saw in the first graph. 

```{r warning=FALSE, message = FALSE}
# Next, we construct the graphs to figure out if if any particular types of
# felonies are causing the increase or decrease in total defendants sentenced

intake_initiation_fel_by_offense <- int_ini_sent %>% 
  group_by(receive_month, Offense_Category_broad) %>% 
  summarise(felonies = n_distinct(CASE_PARTICIPANT_ID)) %>% 
  filter(receive_month > "2012-01-01" & receive_month < "2018-01-01") 
  
intake_initiation_2012 <- intake_initiation_fel_by_offense %>% 
  mutate(receive_month2 = case_when(
    receive_month=="2012-02-01" ~ "Start",
    receive_month=="2017-12-01" ~ "End",
    TRUE ~ "")) %>% 
  group_by(receive_month2) %>% 
  select(-receive_month) %>% 
  filter(receive_month2 == "Start" | receive_month2 == "End") %>% 
  spread(receive_month2,felonies) %>% 
  mutate(change = abs((End-Start)/Start))

int_init_by_off_pct <- left_join(intake_initiation_fel_by_offense, intake_initiation_2012, by="Offense_Category_broad")

int_init_by_off_pct$Offense_Category_broad <- factor(int_init_by_off_pct$Offense_Category_broad, levels = c(
  "Retail Theft", "DUI", "Narcotics","Unlawful Use of Weapon", "Battery","Homicide","MVT",
  "Driving With Revoked License", "Burglary", "Other", "Theft", "Robbery", "Sex Crimes"),ordered=TRUE)
                                                     
plot3 <- int_init_by_off_pct %>% 
  filter(change > 0.66) %>% 
  ggplot(aes(y=felonies, x=receive_month, color=Offense_Category_broad))  +
  geom_point(size=0.2) +
  geom_smooth(size=0.1) +
  facet_wrap(~Offense_Category_broad, scales = "free") +
  scale_color_manual(values=c("blue","blue","blue", 
                              "chocolate3", "chocolate3", "chocolate3")) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(strip.text = element_text(size = rel(0.7))) +
  labs(
    title = "DUIs, Narcotics, and Retail Theft Cases Drive Decrease in \n Number of Cases Sentenced",
    subtitle = "Number of cases sentenced per month by offense type from 2012 to 2017",
    caption = "Source: Cook County State Attorney's Office Data Portal - Initiations, Intake, and Sentences",
    x = "Year",
    y = "Number of Cases Per Month"
  )
ggsave(here("data_output","num_cases_by_offense.pdf"),plot3,device = "pdf",width=8,height=5)
plot3
```


The above graph plots total number of felonies sentenced from 2012 to 2017, faceted by offense type, for the 6 offense types that had the greatest absolute change in number of felonies from 2012 to 2017. Note, the window now starts at 2012 instead of 2011. This is because the Intake data begins in 2011, and since I've only kept the universe of case participants that are in both Intake and Sentences datasets, I shortened the window. Otherwise, cases that take longer to prosecute (e.g. homicide) would exist in very small numbers in 2011 if they were only initiated in 2011, which would artificially deflate the number of homicide convictions.

From the above graph, we see that retail theft, DUIs and narcotics are driving the decline in number of cases sentenced. These trends are partially explained by policy changes instituted by former Governor Rauner and State's Attorney Kim Foxx. Starting in 2011, Chicago had decriminalized possession of small amounts of marijuana, and a statewide bill approved in July 2016 by Governor Rauner further forced police officers to stop arresting those caught with small amounts of drugs. Also in late 2016, SA Kim Foxx announced that retail-theft cases would be considered misdemeanors (not felonies) unless the suspect stole merchandise worth \$1,000 or higher. Until this change, the threshold was \$300. This policy change can clearly be seen in the steep drop in retail-theft cases around 2017.

Having better understood which offense types are driving the decline in number of cases, I turn to looking at which offense types are driving the increase in case length.

```{r warning=FALSE, message = FALSE}

# Want to calculate avg length per case (first collapse by Case participant) and then avg length
# per offense type, broken down into 2 time periods (before and after 2016)
lengths_by_offense_2016 <- int_ini_sent %>% 
  filter(case_length_total < 2000 & case_length_total >0) %>% 
  mutate(after_2016 = ifelse(sentence_month > "2016-01-01","after","before")) %>% 
  group_by(after_2016, Offense_Category_broad, CASE_PARTICIPANT_ID) %>% 
  summarise(case_length_p = mean(case_length_total, na.rm = TRUE))

lengths_by_off_change <-lengths_by_offense_2016 %>% 
  group_by(after_2016, Offense_Category_broad) %>% 
  summarise(mean_length = mean(case_length_p)) %>% 
  spread(after_2016, mean_length) %>% 
  mutate(change = abs((after-before)/before))
  
lengths_by_offense_2016$after_2016 <- factor(lengths_by_offense_2016$after_2016, labels = c("Before 2016", "After 2016"), ordered = TRUE)

plot4<-lengths_by_offense_2016 %>% 
  filter(Offense_Category_broad %in% c("Retail Theft", "Narcotics", "Homicide","DUI","Unlawful Use of Weapon","Sex Crimes")) %>% 
  ggplot(aes(x=case_length_p,fill=after_2016)) +
  geom_density(alpha=0.3) +
  stat_density(aes(x=case_length_p,fill=after_2016),geom="line",position = "identity") +
  facet_wrap(~Offense_Category_broad,ncol=3) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Average case lengths increasing across offense types, \n even as retail theft and narcotics cases decrease",
    subtitle = "Case lengths by offense type from 2011 to 2018",
    caption = "Source: Cook County State Attorney's Office Data Portal - Initiations, Intake, and Sentences",
    x = "Length of Case in Days",
    y = "Density",
    fill = ""
  ) + 
  scale_fill_manual(values = c("blue","chocolate4"))

ggsave(here("data_output","case_length_offense_type.pdf"),plot4,device="pdf",width=8,height=5)
plot4
```

For the figure above, I choose 2016 as the cut off since there were a number of changes in that year that may have impacted the system as a whole, such as the policy changes mentioned above and the election of a new State's Attorney. 

The distribution of case lengths shifts right across all offense types, though it is not clear why. One explanation is that this is a purely mechanical effect - since "lower level" crimes are falling, the crimes left to prosecute are crimes that naturally take longer to prosecute, and this shifts the distribution. However, this explanation does not fully explain why the total number of days spent on cases is $\textit{increasing}$ because there should still be a mechanical decline in time spent if fewer cases are being prosecuted. A second explanation is that there is something about the changes in 2016 causing this increase in case length - new State's Attorneys often comes in with a new set of priorities and procedures, and those could be causing lags.


```{r warning=FALSE, message = FALSE}
# I want to see if there's any interesting variation in case length by District or Judge

judge_count <- int_ini_sent %>% 
  filter(Offense_Category_broad %in% c("Homicide", "Sex Crimes", "DUI", "Retail Theft", "Narcotics","Unlawful Use of Weapon")) %>%
  filter(case_length_total < 2000 & case_length_total > 0) %>% 
  filter(COURT_NAME != "") %>% 
  filter(SENTENCE_JUDGE != "") %>% 
  group_by(COURT_NAME,SENTENCE_JUDGE,CASE_PARTICIPANT_ID) %>%
  summarise(case_length_p = mean(case_length_total, na.rm = TRUE)) %>%
  group_by(COURT_NAME,SENTENCE_JUDGE) %>%
  summarise(total_sentences = n_distinct(CASE_PARTICIPANT_ID), 
            total_case_length = sum(case_length_p, na.rm = TRUE),
            avg_case_length =  mean(case_length_p, na.rm = TRUE))

judge_count %>% 
  filter(COURT_NAME %in% c("District 1 - Chicago","District 2 - Skokie","District 5 - Bridgeview",
                           "District 6 - Markham")) %>% 
  filter(total_sentences > 100 ) %>% 
  #ggplot(aes(x=total_sentences/1000, y=total_case_length/100000)) +
  ggplot(aes(x=total_sentences, y=total_case_length/1000)) +
  geom_jitter(aes(size=avg_case_length,color=COURT_NAME),alpha=0.5, position = "jitter") +
  theme_minimal() +
  scale_y_log10() +
  labs(
    title="Delays in the sentencing not uniformly distributed across judges and districts",
    x = "Total Cases Sentenced",
    y = "Total Number of Days Spent on Cases \n (Thousands of Days)",
    subtitle="Judges from four largest districts sentencing homicides, sex crimes, DUIs, retail thefts, \n narcotics, and unlawful use of weapon cases",
    color = "Circuit Court District",
    size = "Average Case \n Length (Days)",
    caption = "Source: Cook County State Attorney's Office Data Portal - Initiations, Intake, and Sentences"
  ) 


```

Beyond offense types, I am also interested in seeing if the delays in sentencing cases are uniformly distributed across judges and districts. The figure above suggests that they are not. For the same total number of cases sentenced, we see fairly large variation in the total number of days spent on all cases by judges. For example, for judges who sentenced approximately 500 cases, some judges - particularly in courts serving the southern suburbs (District 6) - spent far more days than other judges. Judges from Markham seem to consistently spend more days on cases than their counterparts in Chicago, Skokie, and Bridgeview. This could be signal that more resources are needed in Markham, perhaps, but more work is needed to understand these trends.





